Emulsions Stabilized by Magnetic Ellipsoidal Particles: A Lattice Boltzmann Study

### 2025-04-07 chapters/dissertation-tap.tex
Changes 1. Modified conclusion section in final remarks 2. Changes image sized in structural response of bijels 3. modified abstract and introduction to include structural analysis results

    We first examine the ability of magnetic fields to influence bijel formation by applying a constant field during spinodal decomposition 
    of fluid mixtures containing disc-like, spherical, and rod-like particles. While no significant change in domain size is observed for 
    spherical particles, bijels stabilized by discs and rods show slight increases in domain size and become anisotropic, as measured by 
    tortuosity and directional domain size.Further analysis reveals that the coarsening rates become direction-specific for ellipsoidal particles, 
    suggesting that jamming occurs anisotropically due to particle alignment with the magnetic field. This alignment, governed by the 
    orientation of the magnetic moment relative to the particle's long axis, also affects how particles arrange themselves at the fluid 
    interface.

>>>>

    We first examine the ability of magnetic fields to influence bijel formation by applying a constant field during spinodal decomposition 
    of fluid mixtures containing disc-like, spherical, and rod-like particles. While no significant change in domain size is observed for 
    spherical particles, bijels stabilized by discs and rods show slight increases in domain size and become anisotropic, as measured by 
    tortuosity and directional domain size.Further analysis reveals that the coarsening rates become direction-specific for ellipsoidal particles, 
    suggesting that jamming occurs anisotropically due to particle alignment with the magnetic field. This alignment, governed by the 
    orientation of the magnetic moment relative to the particle's long axis, also affects how particles arrange themselves at the fluid 
    interface. Particle reorientation affects the curvature of the interface as particles with smaller cross sectional area are less disruptive
    to the saddle like interface shape that minimal surfaces possess. Topological examinations of the interface surface, 
    demonstrating that the number of interconnected channels in the system decreased over time and that the average channel sizes obtained follow
    the same trend as that of the domain size.

### 2025-04-07 chapters/final_remarks.tex
Changes 1. Modified conclusion section in final remarks 2. Changes image sized in structural response of bijels 3. modified abstract and introduction to include structural analysis results

Owing to their high surface area to volume ratio, porous materials are seeing a surge in popularity in applications
such as catalysis, battery electrodes and pharmaceuticals. Given their wide variety of applications, identifying 
fabrication techniques that allow access to the various pore length scales is of interest. One such synthesis technique
is emulsion templating which offers a wide variety of accessible microstructures, addition of stimuli response and 
large number of possible systems that can be fabricated. A microstructure that can be fabricated from emulsion templates
is the bicontinuous interfacially jammed emulsion gel (bijel).

Bijels are normally fabricated using Thermally Induced Phase Separation (TIPS). However, TIPS does not allow for continuous
fabrication of bijels. More recent fabrication techniques such as Solvent Transfer Induced Phase Separation (STrIPS) and 
Vapor Induced Phase Separation (VIPS) allow for continuous fabrication and access to various microstructures. Both techniques
require modifications to the initial emulsion mixture to facilitate microstructure adjustments, which affect the final rheological
properties of the system. Decoupling the microstructure and the casting mixture of the bijel would allow for greater flexibility
in the synthesis of the material.

Stimuli response has been used before to modify the microstructure of particle stabilized emulsions. Magnetic fields offer
targeted material response and low applied field strengths necessary for response. Past work investigating the effect of magnetic
stimuli on bijels stabilized with spherical particles yielded little microstructure change. However, anisotropic particles have
also been used to stabilize bijels. Anisotropic particles at interfaces under magnetic fields have been shown to tilt out of 
interfaces. These effects have yet to be captured in bijels stabilized by ellipsoidal particles under magnetic fields.
This study addresses this knowledge gap using a hybrid Molecular-Dynamics multicomponent method Lattice Boltzmann Method.
We split this work into three aims; Aim 1 addressed the microstructure obtained when applying a magnetic field of various strengths 
onto bijels stabilized by ellipsoidal particles, compared to bijels stabilized wih spherical particles. Aim 2 addressed the 
structural response of bijels stabilized by ellipsoidal particles and analyzes the effect of initial order of the particle monolayer
on the structural response observed. Aim 3 addressed the constant shear response of bijels stabilized by ellipsoidal particles
with pre-existing particle order and under magnetic fields.

In Aim 1, bijels stabilized with spherical particles do not respond to the application of a magnetic field. However, bijels stabilized
with ellipsoidal particles have an increase in the average domain size by 3 \%. The microstructure also becomes anisotropic, 
characterized as a change in the directional tortuosity and distribution of channel widths as a function of the distance from the 
interface. The response characterized was particle morphology specific due to the orientation of the particle magnetic moment with 
respect to the direction of the long axis of the particle. This causes the particles to arrange differently on the interface. The 
microstructure anisotropy is caused by direction specific jamming of the particle monolayer originating from particle ordering to 
the magnetic field. The direction specific jamming originates from the orientation of the particles to the magnetic field. The direction
specific jamming also reduces the local curvature of the system as the interface location moves with the particle.
To investigate the role of the particle monolayer in greater detail, we characterize the orientational order to the magnetic field,
angle to the interface and local particle ordering at the interface. The orientational order changes as a function of the applied field 
strength and has particle morphology specific time evolution. When investigating the average interfacial angle of the particles to the 
interface, it is shown how the capillary interactions with the interface differ, leading to the time evolution differences observed.
When analyzing the ordering of particles on the interface, disc-like particles see lowered local ordering as the magnetic field strength is 
increased while rod-like particles see greater local ordering. These properties are attributed to how those particle morphologies prefer 
to orient themselves at interfaces to one another, with disc like particles preferring to stack while rod like particles order side to side 
or end to end.

In Aim 2, bijels stabilized by ellipsoidal particles demonstrate responses to an applied magnetic field. When applying a magnetic field
onto a bijel template simulated with no magnetic fields, we see an average microstructure change of up to 5 \%. This microstructure change
arises from the magnetic field driven re-orientation of particles at the interface. Particle reorientation creates microstructure anisotropy
arising from the alignment of particles to the magnetic field. When analyzing the particle monolayer, we characterize the alignment of particles
to the field direction, average interface angle and the local ordering of the particles. We demonstrate that the average interface angle increases
as the particles reorient to the field before the capillary forces of the interfaces causes the interface to move, until jamming of the monolayer takes
place. The local particle order during this process decreases for oblate particles and increases for prolate particles. We characterize that the local 
order of the particles controls the timescales of response.
When investigating the effect of pre-existing order of the particle monolayer, we show that the structural response of the bijel is dependent upon the
difference between the field applied to make the bijel and the applied field when the field is applied in the same direction. We also characterize that the
local particle monolayer is dependent upon the pre-existing order of the bijel. When switching off the magnetic field onto bijels made with a magnetic field,
we see that the microstructure remains resilient to the removal of the magnetic field. We characterize that while there are differences in the particle monolayer,
these differences manifest as when domain coarsening begins. When characterizing the hysteresis curve of the bijels stabilized with prolate particles,
we see the effect of this.

In Aim 3, the rheology of bijels stabilized by ellipsoidal particles under shear was analyzed, specifically examining the effects of initial microstructure and a 
constant magnetic field on the shear response. Using a Herschel-Bulkley fitting, shear-thinning behavior remains robust although the 
flow index and yield stress are strongly influenced by particle morphology and magnetic field application. Increased particle ordering led to 
lower yield stress in oblate-particle-stabilized bijels and higher yield stress in prolate-particle-stabilized bijels, highlighting the role of particle 
shape in interfacial mechanics. Without a magnetic field, both particle monolayers remained adsorbed at the interface, but when a field was applied, 
oblate particles flipped out of the interface, whereas prolate particles tilted, altering the flow field and impacting non-Newtonian behavior. Additionally, 
coarsening driven by reduced interfacial particle coverage emerged as the dominant mechanism behind domain size evolution in sheared bijels. The anisotropic 
domain growth trends matched expectations in all cases except when the field was removed, indicating that the history of field application significantly influences 
microstructural evolution. These findings provide insight into the interplay between particle morphology, interfacial coverage, and external fields, contributing 
to the understanding of magnetically tunable soft materials and structured emulsions.

@@

This research was supported in part through the use of DARWIN computing system: DARWIN - A Resource for Computational 
and Data-intensive Research at the University of Delaware and in the Delaware Region, Rudolf Eigenmann, Benjamin E. 
Bagozzi, Arthi Jayaraman, William Totten, and Cathy H. Wu, University of Delaware, 2021, URL: https://udspace.udel.edu/handle/19716/29071

>>>>

Porous materials, owing to their high surface-area-to-volume ratio, have gained significant attention in a wide range of 
applications including catalysis, battery electrodes, and pharmaceuticals. Their functionality across diverse sectors has 
driven a growing interest in fabrication techniques capable of accessing and tuning pore structures across multiple length 
scales. Emulsion templating has emerged as a particularly versatile synthesis route, offering access to a broad spectrum of 
microstructures, the potential for stimuli-responsive behavior, and compatibility with a wide variety of material systems. 
Among the structures that can be generated through emulsion templating, the bicontinuous interfacially jammed emulsion gel 
(bijel) stands out for its unique interconnected morphology and potential for functional material design.

Traditional bijel fabrication relies on thermally induced phase separation (TIPS), a method that, while effective, does not 
support continuous production. More recently developed techniques—such as Solvent Transfer Induced Phase Separation (STrIPS) 
and Vapor Induced Phase Separation (VIPS)—enable continuous fabrication and offer greater control over microstructure. However, 
these methods typically require modifications to the initial emulsion formulation, coupling the casting mixture to the resulting 
morphology. Decoupling these parameters would offer significantly greater flexibility in bijel design, allowing for independent 
tuning of material properties and processing conditions.

External stimuli have been previously employed to modulate the microstructure of particle-stabilized emulsions, with magnetic 
fields offering particular promise due to their directionality and the relatively low energy required to induce a response. Prior 
studies have shown that bijels stabilized with spherical particles exhibit limited structural response to magnetic fields. In 
contrast, anisotropic particles such as ellipsoids have demonstrated the capacity to reorient and tilt out of fluid interfaces 
under magnetic influence. However, the effect of magnetic fields on bijels stabilized by ellipsoidal particles remains unexplored.

This study addresses that gap through a series of hybrid simulations using a coupled Molecular Dynamics-Lattice Boltzmann framework. 
We investigated how magnetic fields influence bijel formation and structure when stabilized by ellipsoidal particles, focusing on 
three key aspects: the impact of magnetic field strength on microstructure compared to systems stabilized by spherical particles; 
the structural response of these bijels under post-formation magnetic fields and how this response depends on the initial ordering 
of the interfacial particle monolayer; and the rheological behavior of such systems under constant shear, particularly in relation 
to prior magnetic field exposure and pre-existing interfacial particle alignment.

We first demonstrated that ellipsoidal particles alter the formation dynamics of bijels when subjected to external magnetic 
fields during phase separation. We also explored how varying magnetic flux densities and particle aspect ratios affect structural outcomes. 
In the absence of a magnetic field, ellipsoidal particles 
aligned with the interface over time due to capillary interactions, contributing to slower coarsening and finer domain features compared 
to spherical particles. Upon introducing magnetic fields, particles experienced additional torques that promoted alignment along the field 
direction during formation. This field-induced reorientation improved interfacial packing efficiency and introduced directional cessation
of coarsening, leading to bijels with domain anisotropy. Importantly, these effects were achieved without preventing the formation of a 
stable, bicontinuous structure.

Structural characterization of bijels stabilized by ellipsoids revealed morphological detail across particle shapes and magnetic field strengths. 
Six-fold bond orientational order emerged within the interfacial particle layers, but true crystalline order remained suppressed due to the 
inherent curvature of the interface. Ellipsoidal particles increased the Gaussian curvature of the interface compared 
to spheres, while increasing field strength tended to reduce this curvature by promoting particle alignment due to reductions in the interface
deformation of each particle. This reduction in curvature indicated a more hyperbolic interfacial character.

Topological analysis showed that the number of handles, or interconnected channels, decreased over time as coarsening progressed, following 
a power-law relationship inversely proportional to domain size. After the onset of jamming, this decay slowed considerably. Notably, 
channel size distributions (CSDs) revealed a wide range of channel dimensions but did not show systematic variation with 
field strength, suggesting that magnetic field effects manifest more prominently in local structure and domain anisotropy. Still, the 
average channel size derived from the CSD aligned well with conventional domain measurements, reinforcing the results obtained throughout
the course of this work.

Upon applying magnetic fields post formation, we observed significant structural rearrangements in bijels depending on the initial particle 
configuration. Systems with disordered interfacial layers underwent domain coarsening and anisotropic growth due to field-induced unjamming 
and subsequent re-jamming. Specifically, oblate particles induced perpendicular domain elongation of up to 60\%, while prolate particles led 
to axial coarsening of up to 40\%. These shape-dependent responses were driven by particle rotation and interface realignment under magnetic 
torque, and were quantified using average interface angles and Steinhardt $Q_6$ bond orientational order parameters. Importantly, once the 
magnetic field was removed, the bijel structure largely remained in its field-aligned state. Only minor particle relaxation occurred, 
indicating that capillary forces alone were insufficient to reverse the structural transformations, revealing a hysteresis-like behavior 
in the response to magnetic stimuli.

The degree of structural reorganization was closely linked to the initial state of the particle monolayer. We identified two primary 
response modes. The first is where initially disordered layers unjammed and reorganized under magnetic fields before settling into a new 
jammed configuration, and another where partially ordered particles reoriented to align with the field, modifying domain structure 
without full unjamming. Systems with high initial order showed minimal structural change, suggesting that the dynamic response of bijels 
can be pre-programmed via the interfacial order established during formation.

The rheological behavior of these systems was equally sensitive to particle shape, magnetic field history, and interfacial order. All 
bijels exhibited shear-thinning behavior, which is characteristic of non-Newtonian fluids. However, yield stress and flow 
response varied significantly across systems. Prolate particle-stabilized bijels exhibited higher yield stress and more consistent flow 
behavior with increasing magnetic alignment, attributed to improved interfacial adhesion and stress transfer. Conversely, oblate-stabilized 
systems showed a decrease in yield stress when subjected to the same field history. This contrast was explained by the tendency of oblate 
particles to tilt away from the interface under magnetic alignment, making the interfacial layer more susceptible to shear-induced buckling.

Steinhardt $Q_6$ values and interface angle analyses further supported these interpretations. Higher $Q_6$ values in aligned prolate systems 
indicated robust interfacial ordering and enhanced structural resistance under flow, while oblate systems with elevated interface angles 
demonstrated weaker anchoring and a greater likelihood of particle dislodgment. The impact of particle geometry on interfacial mechanics 
and flow resistance highlights the importance of shape-selective design for applications requiring tunable rheological properties, such as 
drug delivery, 3D printing, or adaptive templating.

This work provides several key insights that significantly advance the current understanding of field-responsive emulsions and structurally 
tunable soft materials. By integrating ellipsoidal particle design with magnetic field stimuli, we demonstrate a new axis of control over 
bijel formation and function. 
Conceptually, our results refine the understanding of interfacial jamming in bijels. The identification of distinct 
unjamming-rejamming mechanisms, modulated by both field strength and particle ordering, reveals that bijel structures are not passively 
determined by initial phase separation dynamics. Instead, they are dynamically reconfigurable under external fields, with the extent of 
reconfiguration governed by the memory of prior structural states. This introduces a paradigm in which bijels can be treated as 
non-equilibrium materials with tunable hysteresis, offering a controllable blend of stability and adaptability.

Our work also implements structural characterization for describing bijel microstructures. While prior studies have focused heavily on global 
measures such as domain size or volume fraction, our use of bond orientational order, curvature metrics, and channel topology creates a more 
nuanced toolkit for evaluating local and global bijel structure. This multiscale perspective provides better predictive capacity for 
tailoring materials with application-specific structural features, such as tortuosity for mass transport or local ordering for mechanical 
performance.

The broader impacts of this work span several disciplines. In materials science, our findings inform the design of smart scaffolds and 
porous networks with tunable permeability or rheology. For instance, magnetic field-tunable tortuosity or anisotropy could 
be leveraged in filtration membranes that dynamically adjust flow pathways or in tissue scaffolds where curvature-sensitive cell adhesion 
is critical. In energy systems, the ability to control interfacial curvature and connectivity may influence electrochemical transport and 
reaction kinetics in battery electrodes or catalytic substrates.

The rheological tunability demonstrated here has direct implications for soft robotics, adaptive damping materials, and controlled release 
systems. The distinction between oblate and prolate behavior under shear and field stresses provides a foundation for selecting particle 
geometries based on the desired flow performance of the end-use material. Furthermore, the discovery that certain configurations can resist 
flow better than others may lead to new materials with programmable viscoelasticity—materials that stiffen or soften in response to field inputs.

@@

This research was supported in part through the use of DARWIN computing system: DARWIN - A Resource for Computational 
and Data-intensive Research at the University of Delaware and in the Delaware Region, Rudolf Eigenmann, Benjamin E. 
Bagozzi, Arthi Jayaraman, William Totten, and Cathy H. Wu, University of Delaware, 2021

### 2025-04-07 chapters/introduction_new.tex
Changes 1. Modified conclusion section in final remarks 2. Changes image sized in structural response of bijels 3. modified abstract and introduction to include structural analysis results

This dissertation investigates the potential of magnetically responsive ellipsoidal particles to enable stimuli-responsive control over bijel microstructure. Specifically, it examines whether magnetic fields 
applied during or after bijel formation can induce meaningful structural changes, and how these changes, in turn, influence the rheological behavior of the resulting material. The central hypothesis is that 
particle reorientation under magnetic fields alters interfacial packing, ultimately reshaping domain morphology. 
Moreover, Carmack and Millet identified that the formation of particle chains was shown to play a critical role in phase separation dynamics. These chains acted as nucleation sites, guiding the development of 
fluid domains and influencing average channel diameter and particle localization at interfaces. These findings suggest that external fields, particularly when coupled with anisotropic particles, offer a 
promising route toward responsive and reconfigurable bijel architectures.
The remainder of this dissertation is organized as follows. Chapter 32 details the physics being modeled and the simulation methods used to model these physics, including the Lattice Boltzmann approach 
and magnetic dipole modeling. Chapter 3 investigates the effects of magnetic fields applied during bijel formation, analyzing their influence on domain structure and particle orientation. Chapter 4 explores 
post-formation responsiveness, specifically examining particle unjamming and reordering phenomena. Chapter 5 assesses the rheological implications of these structural changes, with particular attention to 
yield stress and shear-thinning behavior. Finally, Chapter 6 summarizes the key findings, discusses their implications for the design of tunable porous materials, and outlines potential directions for future 
research.

>>>>

This dissertation investigates the potential of magnetically responsive ellipsoidal particles to enable stimuli-responsive control over bijel microstructure. Specifically, it examines whether magnetic fields 
applied during or after bijel formation can induce meaningful structural changes, and how these changes, in turn, influence the rheological behavior of the resulting material. The central hypothesis is that 
particle reorientation under magnetic fields alters interfacial packing, ultimately reshaping domain morphology. 
Moreover, Carmack and Millet identified that the formation of particle chains was shown to play a critical role in phase separation dynamics. These chains acted as nucleation sites, guiding the development of 
fluid domains and influencing average channel diameter and particle localization at interfaces. These findings suggest that external fields, particularly when coupled with anisotropic particles, offer a 
promising route toward responsive and reconfigurable bijel architectures.
The remainder of this dissertation is organized as follows. Chapter 2 details the physics being modeled and the simulation methods used to model these physics, including the Lattice Boltzmann approach 
and magnetic dipole modeling. Chapter 3 investigates the effects of magnetic fields applied during bijel formation, analyzing their influence on domain structure and particle orientation. 
Chapter 4 investigates the microstructure of bijels stabilized by ellipsoidal particles under magnetic fields in more detail, providing further detail on the influence of the magnetic field on
the coarsening dynamics of the bijel.
Chapter 5 explores 
post-formation responsiveness, specifically examining particle unjamming and reordering phenomena. Chapter 6 assesses the rheological implications of these structural changes, with particular attention to 
yield stress and shear-thinning behavior. Finally, Chapter 7 summarizes the key findings, discusses their implications for the design of tunable porous materials, and outlines potential directions for future 
research.
